---
title: "Data Analysis of English Premier League 2019/20 Season - The Lockdown Games"
output:
  html_document: default
  pdf_document: default
---

## Research topic: Is there a relationship between playing behind closed doors and team performance in football matches during the 2019/2020 season? Performance is classified on the basis of wins, losses, and draws.

#### Project structure: Dataset -> Descriptive Statistics -> Data Preparation & Pre-Processing -> Exploratory Data Analysis -> Modeling -> Recommendations -> References

```{r include=FALSE}
library(ggplot2)
library(tidyverse) 
library(dplyr)
library(viridisLite)
library(gridExtra)
```

## Data-set,Intentions and Approach
After Downloading the data from https://www.kaggle.com/datasets/idoyo92/epl-stats-20192020/data we will get two data set. One about the player statistics and another about team statistics.
As our research question refers to the analysis of team data, my approach is to first have a basic exploration (descriptive statistics, basic exploratory data analysis) of player data set and an in-depth analysis of team data to find new trends. Merging the data set will result in redundancy. For example, in player data a players 'influence'/'creativity'/'threat' adds in his teams xG(expected goals).

```{r}
player <- read.csv("players_1920_fin.csv",TRUE,",")
head(player)
#glimpse(player)
```

The data set has 22,502 observations in 35 variables.<br>

'X': referring the row numbers.<br>
'assists': assists for scoring goals.<br>
'bonus': is fantasy game points.<br>
'bps': is bonus point system which is also fantasy aspect.<br>
'clean_sheet': clean sheet refers to zero goals scored by opponents.<br>
'creativity': Creativity is a statistic that measures a player's ability to produce goal-scoring opportunities for their teammates.<br>
'element': in Fantasy Premier League (FPL), an "element" refers to a player. Each player available for selection in the Fantasy Premier League is categorized as an element.<br>
'fixture': doesn't have any associated value.<br>
'goals_conceded':number of conceded goals.<br>
'goals_scored': number of scored goals.<br>
'influence': Influence measures a player's impact on a match.<br> 
'creativity': Creativity is a statistic that measures a player's ability to produce goal-scoring opportunities for their teammates.<br>
'threat': Threat assesses a player's goal-scoring potential.
'ict_index': culmination of 'influence','creativity' and 'threat'.High ICT Index scores indicate players who are actively involved in the game, contribute offensively, and pose a goal-scoring threat.<br>
'kickoff_time': the time game started.<br>
'minutes': minutes played by the player.<br>
'own_goal': own goal caused by that player.<br>
'penalties_missed': number of penalty missed.<br>
'penalties_saved': number of penalty saved.<br>
'red_cards': number of red cards taken by a player.<br>
'round': number of gameweeks.<br>
'saves': number of saves by goal-keeper.<br>
'team_a_score': teams away score.<br>
'team_h_score': teams home score.<br>
'was_home': home or away?<br>
'yellow_cards': number of yellow cards.<br>
'full': full name of the player.<br>
'team': players team.
'total_points': is the fantasy points they had.<br>
'selected': is how many the player was selected in total including that match.<br>
'transfers_in': how many people took them in their fantasy team around the time period of that match.<br>
'transfers_out': how many people dropped them from their fantasy team around the time period of that match.<br>
'transfers_balance': 'transfers_in'-'transfers_out'.<br>
'value': players fantasy value in millions.<br>
'ppm': fantasy points per match.<br>

## Descriptive Statistics

```{r}
summary(player)

```

## Data Preparation and Pre-Processing.

```{r}
sapply(player, function(x) sum(is.na(x)))
```

There is no missing values in player data set.

### Dropping variables and the intentions behind them.

Most of the variable in player data set are from Fantasy Premier League (www.premierleague.com, n.d.) which may include statistics like how much they cost, how many people owns them, transfers, etc. In other word, irrelevant in our case.

Dropping the first 'X' column which is just row numbers and doesn't have any significant value.<br>
As we have 'ict_index' of a player which was gained by 'influence','creativity', and 'threat.So the later three variables would be redundant.<br>
'bonus': is fantasy game points.<br>
'bps': is bonus point system which is also fantasy aspect.<br>
'fixture': doesn't have any associated value.<br>
'element': in Fantasy Premier League (FPL), an "element" refers to a player. Each player available for selection in the Fantasy Premier League is categorized as an element.<br>
'total_points': is the fantasy points they had.<br>
'selected': is how many the player was selected in total including that match.<br>
'transfers_in': how many people took them in their fantasy team around the time period of that match.<br>
'transfers_out': how many people dropped them from their fantasy team around the time period of that match.<br>
'transfers_balance': 'transfers_in'-'transfers_out'.<br>
'value': players fantasy value in millions.<br>
'ppm': fantasy points per match.<br>

```{r}
#dropping variables using subset function
player = subset(player, select = -c(X,influence, creativity, threat, bonus, bps, fixture, element, total_points, selected, transfers_in, transfers_out, transfers_balance, value, ppm))
```

Some variable name sounds ambiguous. Lets give them appropriate name.

```{r}
# using rename function of dplyr package
player <- player %>%
  rename(Player = full, GameWeek = round)
```

```{r}
# taking some variables and making a data frame for a overall view of a players performance.
player_summary <- player %>%
  group_by(Player) %>%
  summarise(
    Team = unique(team),
    Home_Matches = sum(was_home == TRUE),
    Away_Matches = sum(was_home == FALSE),
    Matches_Played = Home_Matches + Away_Matches,
    Assists = sum(assists),
    Clean_Sheets = sum(clean_sheets),
    Goals_Conceded = sum(goals_conceded),
    Goals_Scored = sum(goals_scored),
    ICT_index = sum(ict_index),
    Minutes_Played = sum(minutes),
    Own_Goals = sum(own_goals),
    Penalties_Missed = sum(penalties_missed),
    Penalties_Saved = sum(penalties_saved),
    Red_Cards = sum(red_cards),
    Yellow_Cards = sum(yellow_cards),
    Saves =  sum(saves)
  )
print(player_summary)
```


```{r}
# player home stats
home_summary <- player %>%
  filter(was_home == TRUE) %>%
  summarise(
    ict_index = sum(ict_index),
    minutes_played = sum(minutes),
    assists = sum(assists),
    goals_scored = sum(goals_scored),
    penalties_missed = sum(penalties_missed),
    red_cards = sum(red_cards),
    yellow_cards = sum(yellow_cards)
  )
print(home_summary)
```

```{r}
away_summary <- player %>%
  filter(was_home == FALSE) %>%
  summarise(
    ict_index = sum(ict_index),
    minutes_played = sum(minutes),
    assists = sum(assists),
    goals_scored = sum(goals_scored),
    penalties_missed = sum(penalties_missed),
    red_cards = sum(red_cards),
    yellow_cards = sum(yellow_cards) 
  )
print(away_summary)
```

These extracted data frame could be used for exploratory data analysis.One research question I want to add with player data set is that:
### Goalkeeprs, Defenders, Midfielders or Strikers, which category of player had better performance? 





The research question will be analysed is:

## Is there a relationship between playing behind closed doors and team performance in football matches during the 2019/2020 season? Performance is classified on the basis of wins, losses, and draws.

I think closed doors refers to no live audience, or home audience, because home fans support, motivation, etc plays and unmeasurable part in the game of football. I will try to analyse team aspects in light of Home and Away matches.If there is significant trends in any side, we could come to some valuable conclusion or insight.

## Introduction of Data-set

```{r}
team <- read.csv("epl2020.csv",TRUE,",")
head(team)
```

Each row represents data about the game. There are 576 observations, so there are 288 games in total. Because one row is for home team data another for away team.For example, First two rows represents Liverpool and Norwich match data.

The data set includes (45 columns) lots of different statistics about the football game: str(team) gives us the datatype of each variable.

X = Indicator for row numbers.<br>
h_a = Teams playing home or away.<br>
xG =  Expected goals.<br>
xGA = Expected goals against.<br>
npxG = Non-penalty Expected goals.<br>
npxGA = Non-penalty expected goals against.<br>
deep,deep allowed = 'deep' is number of final third entries in opponent side and 'deep allowed' is allowed final third entries of opponents.<br>
xpts = Expected points of a team.<br>
npxGD = non-penalty expected goal difference<br>
ppda_cal, allowed ppda = Passes Per Defensive Action.What number of passes was played before a defensive action was taken.<br>
HS = Home Team Shots.<br>
HST = Home Team Shots on Target.<br>
HF = Home Team Fouls Committed.<br>
HC = Home Team Corners.<br>
HY = Home Team Yellow Cards.<br>
HR = Home Team Red Cards.<br>
AS = Away Team Shots.<br>
AST = Away Team Shots on Target.<br>
AC = Away Team Corners.<br>
AF = Away Team Fouls Committed.<br>
AY = Away Team Yellow Cards.<br>
AR = Away Team Red Cards.<br>
HtrgPerc = Shooting Accuracy Percentage.<br>
AtrgPerc = Attack Accuracy Percentage.<br>
scored = Goals scored in that match.<br>
missed = Goals conceded in that match.<br>
results = Match result(target class).<br>
date = Match date.<br>
matchTime = match time.<br>
B365H.x = Betting odds in Home.<br>    
B365D.x = Betting odds for a Draw.<br>      
B365A.x = Betting odds in Away.<br>
pts = Points in that match.<br>
tot_points = match points + previous points.<br>
tot_goal = match goals + previous goals.<br>
round = Match weeks.<br>
wins, draws, loses = Match results.<br>
tot_con = match conceded goals + previous conceded goals.<br>
matchDay = Day of the week.<br>
Referee.x = Match officials.<br>
teamId = team name.<br>

```{r}
summary(team)
```

Summary() provides us the minimum value,value of the 1st quartile (25th percentile), median,3rd quartile (75th percentile), and maximum value of numerical variables. For instance, if we look at the 'xG' variable There are little difference from min value(0.05) to 3rd quartile(1.97), but significant gap between 3rd and max value(6.63).We could assume 75% of the matches had less or equal to 2 expected of goals per game and 25% matches had more than 2 expected goals.We could have ideas about the distributions of variables from the summary.

## Data Preparation and Pre-Processing

```{r}
sapply(team, function(x) sum(is.na(x)))
```

There is no null or missing values in the data set.

## Cleaning the Data Set
First things first, let us try to get rid of redundant variables. 
1. Column 'X' is just numbers for the rows.We can drop it.<br>
2. 'pts' variable is points per match and 'tot_points' is calculated points of past matches + that match.
Lets drop 'tot_points'<br>
3. 'scored' is goals scored per match and 'tot_goal' is overall goals of a team on that game + goals of past matches.Lets drop 'tot_goal'.<br>
4. 'B365H.x','B365D.x','B365A.x' are betting odds so I will drop them.<br>
5. In same way 'missed' is conceded goals per match and 'tot_con' is goals conceded that match + overall goal conceded past matches. I will drop 'tot_con'.<br>
6. 'round' is for tracking each match-week.<br>
7. The 'result' (traget) column is culmination of 'wins', 'draws', and 'loses', so we could drop this three columns.<br>

```{r}
#using select() function to drop variables.
team = subset(team, select = -c(X,B365H.x, B365D.x,B365A.x,tot_goal, tot_con,round, wins, draws, loses))
colnames(team)
```

'date' consists of date+time.We could just transform the variable to date.We have separate variable for match time.

```{r}
team$date <- as.Date(team$date)
print(team$date,5)
```

Let's rename some variables into relevant name.

```{r}
# using rename function of dplyr package
team <- team %>%
  rename(conceded = missed, Home_Away = h_a)
```

Rearranging variables for better analysis.

```{r}
team <- team[, c("teamId","Home_Away","xpts","pts","tot_points","xG","npxG","scored","xGA","npxGA","conceded","npxGD","deep","deep_allowed","ppda_cal","allowed_ppda","HS.x","HST.x","HF.x","HC.x","HY.x","HR.x","AS.x","AST.x","AF.x","AC.x","AY.x","AR.x","HtrgPerc","AtrgPerc","date","matchtime","Referee.x","matchDay","result")]
```


## Exploratory Data Analysis

The exploratory data analysis will be conducted in two parts. The first part will involve analyzing team performance from the perspective of Home and Away games. The second part will focus on assessing the performance of individual teams throughout the season.


### Is there any bias in Home and Away Matches?

```{r}
print(sum(team$Home_Away=="h"))
print(sum(team$Home_Away=="a"))
```

There is exact number of home and away matches.So, we can consider no biases in our analysis for home and away.

### Which days and times are popular for matches?

```{r}
matchday <- ggplot(team) +
  geom_bar(mapping = aes(y = matchDay, fill = factor(matchDay))) +
  labs(title = "Matchday Frequencies", x = "", y = "") +
  scale_fill_viridis_d()

matchtime <- ggplot(team) + 
  geom_bar(mapping = aes(x= matchtime, fill=factor(matchtime))) +
  labs(title = "Matchtime Frequencies", x = "", y = "") +
  scale_fill_viridis_d()


# Arrange the visualizations side by side
grid.arrange(matchday,matchtime,ncol= 2)
```

The majority of matches were scheduled for Saturdays at 3 P.M, indicating a substantial contrast between this day and time to the rest of the week.

### Are there any favoritism in match officials?

```{r}
ggplot(team) + 
  geom_bar(mapping = aes(y = reorder(Referee.x, table(Referee.x)[Referee.x]), fill = Referee.x)) +
  scale_fill_viridis_d() +
  labs(x = "Count", y = "Referee")

```

There is no evident favoritism in match officiating observed across the 22 referees. Specifically, eight officials officiated over 40 or more matches.


### Teams Home and Away performance.Does it influence the game?

Generally, away team performs poorly (this could be because of away fans and habit of not playing in away stadiums) and It would be interesting to see the comparison between away and home performance in closed doors.

Let us make a basic summary for both home and away matches, so that, we could view further inside the data.

### Home statistics of teams
```{r}
# Summarized multiple statistics for each team playing home
team_home_summary <- team %>%
  group_by(teamId) %>%
  filter(Home_Away == "h") %>%
  summarise(
    Matches = sum(Home_Away =="h"),
    Wins = sum(ifelse(result == "w", 1, 0)),
    Draws = sum(ifelse(result == "d", 1, 0)),
    Loses = sum(ifelse(result == "l", 1, 0)),
    Goal_Scored = sum(scored),
    Goal_Conceded =  sum(conceded),
    Goal_Differnce = Goal_Scored - Goal_Conceded,
    Points = sum(pts),
    Success = paste0(as.integer((Points / (Matches * 3)) * 100),"%")
 
  ) %>%
    arrange(desc(Points)) # ordering by points

print(team_home_summary)
```

*Home Games*:
*Liverpool's Perfect Record*:Liverpool won all 15 of their home games, achieving a 100% success rate.
They scored the most goals (40) and had the highest goal difference (+28).

*Strong Home Performances*: Leicester, Manchester City, and Manchester United performed well at home with win rates over 60%. Manchester City's goal difference (+21) was particularly impressive given they played fewer matches (13) compared to others (15).

*Mixed Results for Mid-Table Teams*: Teams like Arsenal, Tottenham, and Everton had solid home records but still showed some vulnerability, with win rates around 50-60%.
Crystal Palace and Wolves had identical points (22) and similar goal differences, highlighting their competitive nature.

*Struggles at Home*: Teams like Norwich, West Ham, and Southampton struggled at home with win rates below 35%.
Southampton had the poorest goal difference at home (-15), reflecting defensive issues.


### Away statistics of teams
```{r}
# Summarized multiple statistics for each team playing away
team_away_summary <- team %>%
  group_by(teamId) %>%
  filter(Home_Away  == "a") %>%
  summarise(
    Matches = sum(Home_Away =="a"),
    Wins = sum(ifelse(result == "w", 1, 0)),
    Draws = sum(ifelse(result == "d", 1, 0)),
    Loses = sum(ifelse(result == "l", 1, 0)),
    Goal_Scored = sum(scored),
    Goal_Conceded =  sum(conceded),
    Goal_Differnce = Goal_Scored - Goal_Conceded,
    Points = sum(pts),
    Success = paste0(as.integer((Points / (Matches * 3)) * 100),"%")
    
  ) %>%
    arrange(desc(Points)) # ordering by points

print(team_away_summary)
```

*Away Games*:
*Liverpool Dominance Continues*: Liverpool also excelled away from home, winning 12 out of 14 matches with an 88% success rate. They maintained a strong goal difference (+17).

*Manchester City's Away Performance*: Manchester City had a good away record with 9 wins out of 15 matches.
They were the highest-scoring team away from home (35 goals).

*Inconsistency Among Top Teams*: Chelsea and Leicester had respectable away records but were less dominant compared to their home performances. Both teams showed competitive goal differences (+5 and +15, respectively).

*Surprising Away Strength*: Southampton performed better away (47% success) compared to their home record.
Sheffield United showed strong resilience away with a high number of draws (7 out of 13).

*Poor Away Records*: Norwich, Aston Villa, and Bournemouth had very poor away performances, with Norwich securing only 6 points and having a goal difference of -20.
Teams like Watford and Brighton also struggled significantly, with low success rates and negative goal differences.

*Summary*:
1. Liverpool was dominant both at home and away, with perfect home results and near-perfect away results.<br>
2. Manchester City and Chelsea performed strongly but were more consistent at home.<br>
3. Southampton and Sheffield United showed relatively better resilience in away games.<br>
4. Norwich, Aston Villa, and Bournemouth struggled significantly in both home and away games, particularly away, where they had very low success rates and poor goal differences.<br>


Lets have a quick look on teams away and home points with pie chart.

```{r}
#pie chart to see all teams home points.
pie_home <- ggplot(team_home_summary, aes(x = teamId, y = Points, fill = teamId)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("x") +
  labs(title = "Pie Chart of Home Points",x="Teams",y="Points")

print(pie_home)
```
```{r}
pie_away <- ggplot(team_away_summary, aes(x = teamId, y = Points, fill = teamId)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("x") +
  labs(title = "Pie Chart of Away Points",x="Teams",y="Points")

print(pie_away)

```

Both charts gives us a idea that very few team does well in away stadiums. Let's have a look at home and away values for some statistics.

```{r}
library(gridExtra)

# Home vs Away Points
points_data <- c(
  sum(team$pts[team$Home_Away == "h"]),
  sum(team$pts[team$Home_Away == "a"])
)
  points <- ggplot() +
  geom_bar(aes(x = "", y = points_data, fill = c("Home", "Away")), stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Points in Home vs Away", x = "", y = "") +
  scale_fill_viridis_d() +
  geom_text(aes(x = "", y = points_data, label = points_data), position = position_stack(vjust = 0.5), color = "white")

# Home vs Away Goals Scored
goals_data <- c(
  sum(team$scored[team$Home_Away == "h"]),
  sum(team$scored[team$Home_Away == "a"])
)
  scored <- ggplot() +
  geom_bar(aes(x = "", y = goals_data, fill = c("Home", "Away")), stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Goals Scored Home vs Away", x = "", y = "") +
  scale_fill_manual(values = c("Home" = "#2ca02c", "Away" = "#9467bd")) +
  geom_text(aes(x = "", y = goals_data, label = goals_data), position = position_stack(vjust = 0.5),color="white")


# Home vs Away Goals conceded
goals_conc <- c(
  sum(team$conceded[team$Home_Away=="h"]),
  sum(team$conceded[team$Home_Away=="a"])
)
  conceded <- ggplot() +
  geom_bar(aes(x = "", y = goals_conc, fill = c("Home", "Away")), stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Goals Conceded Home vs Away", x = "", y = "") +
  scale_fill_manual(values = c("Home" = "black", "Away" = "cyan3")) +
  geom_text(aes(x = "", y = goals_conc, label = goals_conc), position = position_stack(vjust = 0.5),color="white")
  
# Home vs Away xG
xG_data <- c(
    sum(team$xG[team$Home_Away == "h"]),
    sum(team$xG[team$Home_Away == "a"])
  )
  xG_data<- round(xG_data, 0) # round the decimel numbers 

  Expected_Goals <- ggplot() +
  geom_bar(aes(x = "", y = xG_data, fill = c("Home", "Away")), stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Expected Goals Home vs Away", x = "", y = "") +
  geom_text(aes(x = "", y = xG_data, label = xG_data), position = position_stack(vjust = 0.5),color="white")
    

# Arrange the two pie charts side by side
grid.arrange(points, scored,conceded,Expected_Goals, ncol = 2)

```


## Some not so straightforward statistics of the home and away teams.

```{r}

# Get the data from Home stats
home_stats <- c(
  HST = sum(team$HST.x[team$Home_Away == "h"]),
  HF = sum(team$HF.x[team$Home_Away == "h"]),
  HC = sum(team$HC.x[team$Home_Away == "h"]),
  HY = sum(team$HY.x[team$Home_Away == "h"]),
  HR = sum(team$HR.x[team$Home_Away == "h"])
)

# Get the data from Away stats
away_stats <- c(
  
  AST = sum(team$AST.x[team$Home_Away == "a"]),
  AF = sum(team$AF.x[team$Home_Away == "a"]),
  AC = sum(team$AC.x[team$Home_Away == "a"]),
  AY = sum(team$AY.x[team$Home_Away == "a"]),
  AR = sum(team$AR.x[team$Home_Away == "a"])
)

categories <-c (
  "Shoots on target","Fouls","Corners","Yellow Cards","Red Cards"
)

# Combine data into a data-frame
stats <- data.frame(
  Category = categories,
  Home = home_stats,
  Away = away_stats
)

# Load the necessary library
library(ggplot2)
library(ggrepel)
library(tidyr)

# Melt the dataframe for ggplot2 using tidyr
melted_stats <- stats %>%
  gather(key = "Variable", value = "Value", -Category)

# Plot the data
ggplot(melted_stats, aes(x = Category, y = Value, color = Variable, group = Variable)) +
  geom_line() +
  geom_point() +
  geom_text_repel(aes(label = Value), size = 3, nudge_x = 0.2, nudge_y = 0.2, box.padding = 0.5, 
                  point.padding = 0.5, max.overlaps = Inf) +
  labs(title = "Comparison of Home and Away Stats",
       x = "Category",
       y = "Values",
       color = "Legend") +
  theme_minimal()

```


## Analysis on teams

### Which side was better defensively? Home or Away?

```{r}
conceded_all <- team %>%
  group_by(teamId, Home_Away) %>%
  summarise(total_conceded = sum(conceded)) %>%
  ungroup()

ggplot(conceded_all, aes(x = Home_Away, y = total_conceded, fill = Home_Away)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = total_conceded), vjust = 1.5, color = "white") +  # Add text labels on bars
  labs(x = "Home/Away", y = "Goals Conceded", title = "Goals Conceded for Home and Away Matches") +
  scale_fill_manual(values = c("h" = "#2ca02c", "a" = "#9467bd")) +
  facet_wrap(~teamId) +
  theme_minimal()

```
*Key insights*:

Teams like *Aston Villa* *Everton* and *Newcastle United* conceded significantly more goals in away matches than at home.
Other teams like *Wolves* and *Norwich* show almost equal performance in terms of goals conceded, regardless of playing home or away.
Teams like *Southumpton*, *Arsenal*, *Liverpool* conceded more goals in home than away.

Liverpool stands out for conceding very few goals, especially in away matches.


### Quick look to all teams performance (wins,draws,loses).

```{r}
# Every teams results(win, lose , draw)

ggplot(team, aes(y = teamId, fill = result)) +
  geom_bar(position = "stack", stat = "count") +
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(vjust = 0.5), color = "white") +
  scale_fill_viridis_d() + scale_fill_manual(values = c("w" = "#2ca02c", "d" = "grey","l"="#9467bd"))
  labs(y = "Teams", x = "Count of Results", title = "Results for Each Team")

```

Key observations:

1. *Liverpool* stands out with the most wins (27) and very few losses and draws, indicating a strong performance.
2. *Man City* and *Leicester* also have a high number of wins (18 and 16, respectively), though with more draws and losses than Liverpool.
3. Teams like *Norwich* and *West Ham* have more losses than wins, suggesting a less successful season.
4. *Arsenal*, *Newcastle*, *Brighton* has a balanced number of wins, losses, and draws, indicating a season of mixed outcomes.


### Overall points timeline of teams

```{r}
library(ggplot2)
library(gganimate)
library(babynames)
library(hrbrthemes)

# Create the animated line graph
p <- ggplot(team, aes(x = date, y = tot_points, group = teamId, color = teamId)) +
  geom_line() +
  geom_point() +
  geom_text(aes(label = teamId), vjust = -1) +
  labs(title = 'Points over Time by Team',
       x = 'Date',
       y = 'Points') +
  theme_minimal() +
  transition_reveal(date)

# Render and save the animation
animate(p, nframes = 100, fps = 10, renderer = gifski_renderer("points_timeline.gif"))

# View the animation
animate(p, nframes = 100, fps = 10)

```



# Modeling

Let's convert the columns as numeric.
```{r}
# converting the datatypes to numeric
team$teamId <- as.numeric(as.factor(team$teamId))
team$Home_Away <- as.numeric(as.factor(team$Home_Away))
team$date <- as.numeric(as.factor(team$date))
team$matchDay <- as.numeric(as.factor(team$matchDay))
team$Referee.x <- as.numeric(as.factor(team$matchDay))
team$result <- as.numeric(as.factor(team$result))

```


```{r}

library(corrplot)

corrplot(cor(team), method="square") 
cor(team)
```

#### Significant Variables with 'result':

xG: 0.35
deep: 0.20
scored: 0.48
pts: 0.68
xGA: -0.18
conceded: -0.25
allowed_ppda: 0.11
HST.x: 0.11
HY.x: -0.10

As we are not determining only winning but also draw and lose, we will be using xGA,npxGA,conceded,HY.x(Home Yellow cards).HST.x refers to Home Shoot on Target which directly attributes to expected goals(xG).
Interesting to see that there are few high correlation between the predictors:
xG,npxG = 0.96.<br>
xGA,npxGA = 0.96.<br>
xpts,npxGD = 0.94.<br>
Which could leads to Multicollinearity. If we notice,higher non-penalty expected goal difference (npxGD) means higher expected points(xpts).Same goes for xGA and npxGA. On the other hand, xG and npXG are basically same.


### Decision Tree

Decision Tree works very well with multiple-class classification.

```{r}
set.seed(123) # to generate the same result every time.

#train test
train_indices <- sample(2,nrow(team),replace = T,prob = c(0.8,0.2))
train_data <- team[train_indices==1, ]
test_data <- team[-train_indices==2, ]

library(rpart)

#fitting model
DTmodel <- rpart(result ~ xG+ xGA+scored+conceded+xpts+deep+pts+ppda_cal+allowed_ppda,HST.x,HY.x,data = team,method = "class")

# Make predictions on the test set
predictions <- predict(DTmodel, newdata = test_data, type = "class")

library(rpart.plot)
rpart.plot(DTmodel)
```

The decision tree primarily relies on *xGA* (Expected Goals Against) and *xG* (Expected Goals) to classify the result into one of three classes.
Class Predictions:
If xGA < 1.4, the result is predicted to be class 3 (53%).
If xGA >= 1.4 and xG >= 1.3, the result is predicted to be class 1 (1%).
If xGA >= 1.4 and xG < 1.3, the result is predicted to be class 2 (45%).

### Neural Networks

Neural Network is inspired from human brain which is consists of input , hidden layer and output. It works well with multiple classification(www.datacamp.com, n.d.).

```{r}
library(neuralnet)

set.seed(123)

#train test
train_indices <- sample(1:nrow(team), 0.8 * nrow(team))
train_data <- team[train_indices, ]
test_data <- team[-train_indices, ]


nn_model <- neuralnet(result ~ xG+ xGA+scored+conceded+xpts+deep+pts+ppda_cal+allowed_ppda+HST.x+HY.x, data = train_data, hidden = c(2)) # 2 hidden layer

# Make predictions 
nn_predictions <- predict(nn_model, newdata = test_data)

# Convert predictions to class labels
nn_predicted_classes <- ifelse(nn_predictions > 2, 3, 1)

# Evaluate the performance
conf_mat_nn <- table(nn_predicted_classes, test_data$result)

# Calculate accuracy, precision, recall, and F1 score
accuracy_nn <- sum(diag(conf_mat_nn)) / sum(conf_mat_nn)
precision_nn <- diag(conf_mat_nn) / rowSums(conf_mat_nn)
recall_nn <- diag(conf_mat_nn) / colSums(conf_mat_nn)
f1_nn <- 2 * (precision_nn * recall_nn) / (precision_nn + recall_nn)

# Print the results
print("\n")
cat("Neural Network Accuracy:", accuracy_nn, "\n")
cat("Neural Network Precision:", precision_nn, "\n")
cat("Neural Network Recall:", recall_nn, "\n")
cat("Neural Network F1 Score:", f1_nn, "\n")

```
*Accuracy (28.45%)*: The model's overall accuracy is quite low, suggesting it often makes incorrect predictions.

*Precision*: The precision for one of the classes is 0, meaning the model does not correctly identify any instances of that class, which is problematic.

*Recall*: The model has perfect recall (1.0) for one class, meaning it correctly identifies all true positives of that class, but has poor recall for the other classes.

*F1 Score*: The F1 scores are inconsistent, with one class showing a decent score (0.62) while another is undefined (NaN), indicating significant imbalance in how the model handles different classes.

The model is heavily biased towards one class, leading to poor performance for other classes. This suggests the need for further tuning, potentially addressing class imbalances or improving model complexity.


## Recommendations

From our models we cannot say that there is a evident connection between playing behind closed doors and team performance in football matches conducted close doors for Covid-19 during the 2019/2020 season. Even though happening close doors, the home and away stats didn't changes that much, in other words, home matches got favorable numbers. We could comment in teams Performance in general which is classified on the basis of wins, losses, and draws. Further, data analysis shows us some interesting insights but not a confident showcase to say there is a relation. The relation was more non-technical which we cannot put in numbers. Interesting thing we can do is investigate previous 2/3 years 
team performace and compare them to this season which could lead to interesting insights.


## References
www.premierleague.com. (n.d.). FPL basics: What is FPL? [online] Available at: https://www.premierleague.com/news/2173986.

www.datacamp.com. (n.d.). ANN (Artificial Neural Network) Models in R: Code & Examples on How to Build Your NN. [online] Available at: https://www.datacamp.com/tutorial/neural-network-models-r.

